Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change

### 2021-01-15 Manuscript/HI_Model_ms.Rmd
updated manuscript files

---
title: "Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change"
bibliography: HI_Model.bib
csl: iop.csl
fontsize: 12pt
geometry: margin=1in
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
output: 
  pdf_document: 
    latex_engine: xelatex
urlcolor: blue
nocite: |
  @jacobi_map_2017
---

```{r setup, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo = FALSE)
```
Paul C. Selmants^1,6^, Benjamin M. Sleeter^2^,  Jinxun Liu^1^, Tamara S. Wilson^1^,  
Clay Trauernicht^3^, Abby G. Frazier^4^, Gregory P. Asner^5^

**Affiliations:**  
^1^U.S. Geological Survey, Moffett Field, CA, USA   
^2^U.S. Geological Survey, Seattle, WA, USA  
^3^University of Hawai`i at Mānoa, Honolulu, HI, USA  
^4^The East-West Center, Honolulu, HI, USA  
^5^Arizona State University, Tempe, AZ, USA  
^6^Author to whom correspondence should be addressed

@@

## Abstract
The State of Hawai`i passed legislation to be carbon neutral by 2045, a goal that will partly depend on carbon sequestration by terrestrial ecosystems. However, there is considerable uncertainty surrounding the future direction and magnitude of the land carbon sink in the Hawaiian Islands. We used simulation modeling to assess how projected future changes in climate and land use will influence ecosystem carbon balance in the Hawaiian Islands under all combinations of two radiative forcing scenarios (RCPs 4.5 and 8.5) and two land-use scenarios (low and high) over a 90-year timespan from 2010-2100. Collectively, terrestrial ecosystems of the Hawaiian Islands acted as a net carbon sink under low radiative forcing (RCP 4.5) for the entire 90-year simulation period, with low land-use change further enhancing carbon sink strength. In contrast, Hawaiian terrestrial ecosystems transitioned from a net sink to a net source of CO~2~ to the atmosphere under high radiative forcing (RCP 8.5), with high land-use accelerating this transition and exacerbating net carbon loss. A sensitivity test of the CO~2~ fertilization effect on plant productivity revealed it to be a major source of uncertainty in projections of ecosystem carbon balance. Reconciling this uncertainty in how net photosynthesis will respond to rising atmospheric CO~2~ will be essential to realistically constrain simulation models used to evaluate the effectiveness of ecosystem‐based climate mitigation strategies.

@@

The State of Hawai`i exemplifies the challenges of projecting future interactive effects of climate and land use change on ecosystem carbon balance. The main Hawaiian Islands are a complex mosaic of natural and human-dominated landscapes overlain by steep climate gradients across relatively short distances. Mean annual temperature ranges from ~4-24° C [@giambelluca_evapotranspiration_2014] and mean annual rainfall ranges from ~180-9500 mm [@giambelluca_online_2013], with minimal seasonal variation in climate. Temperatures have risen rapidly in the Hawaiian Islands since the mid 1970's [@giambelluca_secular_2008] and a long-term drying trend has persisted since the early 1920s [@frazier_spatial_2017], resulting in reduced forest biomass and productivity [@barbosa_effects_2017]. These same drying and warming trends have increased the frequency and intensity of wildland fire [@trauernicht_contemporary_2015; @trauernicht_vegetationrainfall_2019] with predictable negative effects on ecosystem carbon balance [@selmants_baseline_2017]. Ecosystem carbon stocks across the main Hawaiian Islands have also been strongly influenced by the legacy of past land-use change [@asner_high-resolution_2011]. Thousands of hectares of land were deforested beginning in the late 19th century to clear land for sugar plantations and cattle pasture [@cuddihy_alteration_1990]. Beginning in the latter half of the 20th century, this agricultural land has been steadily converted to urban areas, commercial forestry plantations, or simply abandoned and colonized by non-native grass species [@suryanata_diversified_2009; @perroy_evolving_2016].

Hawaii was the first U.S. state to enact legislation supporting commitments and goals of the Paris climate agreement. Specifically, Hawaii passed Act

@@

## Methods
We used the Land Use and Carbon Scenario Simulator (LUCAS), an integrated landscape change and carbon gain-loss model, to project changes in ecosystem carbon balance for the seven main Hawaiian Islands under all combinations of two land-use scenarios (low and high) and two radiative forcing scenarios (RCP 4.5 and RCP 8.5). We also developed a separate set of scenarios to test model sensitivity to different levels of a CO~2~ fertilization effect (CFE). The landscape change portion of LUCAS is a state-and-transition model that applies a Monte Carlo approach to track the state type and age of each simulation cell in response to a pre-determined set of transitions [@daniel_state-and-transition_2016]. The carbon gain-loss portion tracks carbon stocks within each simulation cell over time as continuous state variables, along with a pre-defined set of continuous flows specifying rates of change in stock levels over time [@daniel_integrating_2018; @sleeter_effects_2019]. We parameterized the Hawai`i LUCAS model to estimate annual changes in carbon stocks and fluxes in response to land use, land use change, wildland fire, and long-term climate variability for the time period 2010-2100.

@@

 Transition targets were based on historical trends of land use change in the Hawaiian Islands from 1992-2011 [@noaa_coastal_2020] and on population projections for the State of Hawai`i [@kim_population_2018]. For the high land-use scenario, transition rates for each timestep and Monte Carlo realization were sampled from uniform distributions bounded by the median and maximum historical rates of agricultural contraction, agricultural expansion, and urbanization for each island. For the low land-use scenario, rates of agricultural contraction and expansion were sampled from uniform distributions bounded by zero and the minimum historical rates for each island. Urbanization rates in the low land-use scenario were based on island-level population estimates and projections at five year intervals from 2010-2045 [@kim_population_2018]. We converted population projections into urbanization transition targets following @sleeter_future_2017 by calculating population density for each island and then projecting future developed area based on the five-year incremental change in island population. The spatial extent of agricultural contraction, agricultural expansion, and urbanization was constrained in both land-use scenarios based on existing zoning maps [@daniel_state-and-transition_2016]. Transition targets for tree plantation harvest were set at ~75% of recent historical rates in the high land-use scenario and ~40% of recent historical rates in the low land-use scenario [@daniel_state-and-transition_2016]. In both land-use scenarios, approximately 60% of tree plantation harvests were replacement harvests resulting in conversion to agriculture. The remaining 40% were rotation harvests replanted to *Eucalyptus* spp.

@@

 Initial carbon stocks and baseline carbon flows were estimated based on the moisture zone (figure S1), state type, and age of each simulation cell using a lookup table derived from the Integrated Biosphere Simulator [IBIS; @foley_integrated_1996; @liu_critical_2020], a process-based dynamic global vegetation model. We initiated IBIS with minimal vegetation and simulated forward for 110 years using 30-year climate normals for the Hawaiian Islands [@giambelluca_online_2013; @giambelluca_evapotranspiration_2014]. We calibrated IBIS carbon stocks with statewide gridded datasets of soil organic carbon [@ssurgo_2016] and forest aboveground live biomass [@asner_rapid_2016]. We also calibrated gross photosynthesis in IBIS using a Hawai`i-specific gridded dataset derived from MODIS satellite imagery [@kimball_evaluating_2017].

@@

```{r fig.align = 'center', out.width = "99%", fig.cap = "Sensitivity of projected changes in total ecosystem carbon storage (a) and mean annual net biome productivity (b) to different rates of carbon dioxide fertilization in the seven main Hawaiian Islands under the RCP 8.5 radiative forcing scenario and high land use scenario. The carbon dioxide fertilization effect (CFE) is the percent change in net primary productivity (NPP) for every 100 ppm increase in atmospheric carbon dioxide. The CFE for all rates is capped at 600 ppm, which is achieved around the year 2060. Solid lines in (a) indicate the mean total ecosystem carbon storage across 30 Monte Carlo realizations for each CFE rate, with shaded areas indicating the minimum and maximum range of Monte Carlo realizations. Solid circles in (b) represent mean annual net biome productivity averaged across all years and Monte Carlo realizations for each CFE rate, with vertical lines indicating the standard deviation of the mean. The dashed horizontal line in (b) represents the boundary between ecosystems acting as a net carbon sink (positive NBP values) and a net carbon source (negative NBP values)."}
knitr::include_graphics("./fig_images/fig6_cfe.png") 
```

@@

This study was funded by the U.S. Geological Survey Biological Carbon Sequestration Program. Thanks to Leonardo Frid and Colin Daniel of ApexRMS for assistance with SyncroSim software and to Nicholas Koch of Forest Solutions, Inc. for information on eucalyptus harvesting in Hawaii. Thanks also to Christian Giardina and Zhiliang Zhu for making this project possible. Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.

>>>>

---
title: "Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change"
bibliography: HI_Model.bib
csl: iop.csl
geometry: margin=1in
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
output: 
  pdf_document: 
    latex_engine: xelatex
mainfont: Times New Roman
fontsize: 12pt
urlcolor: blue
nocite: |
  @jacobi_map_2017
---

```{r setup, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo = FALSE)
```
\centering
Paul C. Selmants^1,6^, Benjamin M. Sleeter^2^,  Jinxun Liu^1^, Tamara S. Wilson^1^,

Clay Trauernicht^3^, Abby G. Frazier^4^, Gregory P. Asner^5^

\raggedright

**Affiliations:**  
^1^U.S. Geological Survey, Moffett Field, CA, USA     
^2^U.S. Geological Survey, Seattle, WA, USA    
^3^University of Hawaiʻi at Mānoa, Honolulu, HI, USA  
^4^The East-West Center, Honolulu, HI, USA    
^5^Arizona State University, Tempe, AZ, USA    
^6^Author to whom correspondence should be addressed

@@

## Abstract
The State of Hawaiʻi passed legislation to be carbon neutral by 2045, a goal that will partly depend on carbon sequestration by terrestrial ecosystems. However, there is considerable uncertainty surrounding the future direction and magnitude of the land carbon sink in the Hawaiian Islands. We used simulation modeling to assess how projected future changes in climate and land use will influence ecosystem carbon balance in the Hawaiian Islands under all combinations of two radiative forcing scenarios (RCPs 4.5 and 8.5) and two land-use scenarios (low and high) over a 90-year timespan from 2010-2100. Collectively, terrestrial ecosystems of the Hawaiian Islands acted as a net carbon sink under low radiative forcing (RCP 4.5) for the entire 90-year simulation period, with low land-use change further enhancing carbon sink strength. In contrast, Hawaiian terrestrial ecosystems transitioned from a net sink to a net source of CO~2~ to the atmosphere under high radiative forcing (RCP 8.5), with high land-use accelerating this transition and exacerbating net carbon loss. A sensitivity test of the CO~2~ fertilization effect on plant productivity revealed it to be a major source of uncertainty in projections of ecosystem carbon balance. Reconciling this uncertainty in how net photosynthesis will respond to rising atmospheric CO~2~ will be essential to realistically constrain simulation models used to evaluate the effectiveness of ecosystem-based climate mitigation strategies.

@@

The State of Hawaiʻi exemplifies the challenges of projecting future interactive effects of climate and land use change on ecosystem carbon balance. The main Hawaiian Islands are a complex mosaic of natural and human-dominated landscapes overlain by steep climate gradients across relatively short distances. Mean annual temperature ranges from ~4-24° C [@giambelluca_evapotranspiration_2014] and mean annual rainfall ranges from ~180-9500 mm [@giambelluca_online_2013], with minimal seasonal variation in climate. Temperatures have risen rapidly in the Hawaiian Islands since the mid 1970's [@giambelluca_secular_2008] and a long-term drying trend has persisted since the early 1920s [@frazier_spatial_2017], resulting in reduced forest biomass and productivity [@barbosa_effects_2017]. These same drying and warming trends have increased the frequency and intensity of wildland fire [@trauernicht_contemporary_2015; @trauernicht_vegetationrainfall_2019] with predictable negative effects on ecosystem carbon balance [@selmants_baseline_2017]. Ecosystem carbon stocks across the main Hawaiian Islands have also been strongly influenced by the legacy of past land-use change [@asner_high-resolution_2011]. Thousands of hectares of land were deforested beginning in the late 19th century to clear land for sugar plantations and cattle pasture [@cuddihy_alteration_1990]. Beginning in the latter half of the 20th century, this agricultural land has been steadily converted to urban areas, commercial forestry plantations, or simply abandoned and colonized by non-native grass species [@suryanata_diversified_2009; @perroy_evolving_2016].

Hawaiʻi was the first U.S. state to enact legislation supporting commitments and goals of the Paris climate agreement. Specifically, Hawaiʻi passed Act

@@

## Methods
We used the Land Use and Carbon Scenario Simulator (LUCAS), an integrated landscape change and carbon gain-loss model, to project changes in ecosystem carbon balance for the seven main Hawaiian Islands under all combinations of two land-use scenarios (low and high) and two radiative forcing scenarios (RCP 4.5 and RCP 8.5). We also developed a separate set of scenarios to test model sensitivity to different levels of a CO~2~ fertilization effect (CFE). The landscape change portion of LUCAS is a state-and-transition model that applies a Monte Carlo approach to track the state type and age of each simulation cell in response to a pre-determined set of transitions [@daniel_state-and-transition_2016]. The carbon gain-loss portion tracks carbon stocks within each simulation cell over time as continuous state variables, along with a pre-defined set of continuous flows specifying rates of change in stock levels over time [@daniel_integrating_2018; @sleeter_effects_2019]. We parameterized the Hawai"i LUCAS model to estimate annual changes in carbon stocks and fluxes in response to land use, land use change, wildland fire, and long-term climate variability for the time period 2010-2100.

@@

 Transition targets were based on historical trends of land use change in the Hawaiian Islands from 1992-2011 [@noaa_coastal_2020] and on population projections for the State of Hawaiʻi [@kim_population_2018]. For the high land-use scenario, transition rates for each timestep and Monte Carlo realization were sampled from uniform distributions bounded by the median and maximum historical rates of agricultural contraction, agricultural expansion, and urbanization for each island. For the low land-use scenario, rates of agricultural contraction and expansion were sampled from uniform distributions bounded by zero and the minimum historical rates for each island. Urbanization rates in the low land-use scenario were based on island-level population estimates and projections at five year intervals from 2010-2045 [@kim_population_2018]. We converted population projections into urbanization transition targets following @sleeter_future_2017 by calculating population density for each island and then projecting future developed area based on the five-year incremental change in island population. The spatial extent of agricultural contraction, agricultural expansion, and urbanization was constrained in both land-use scenarios based on existing zoning maps [@daniel_state-and-transition_2016]. Transition targets for tree plantation harvest were set at ~75% of recent historical rates in the high land-use scenario and ~40% of recent historical rates in the low land-use scenario [@daniel_state-and-transition_2016]. In both land-use scenarios, approximately 60% of tree plantation harvests were replacement harvests resulting in conversion to agriculture. The remaining 40% were rotation harvests replanted to *Eucalyptus* spp.

@@

 Initial carbon stocks and baseline carbon flows were estimated based on the moisture zone (figure S1), state type, and age of each simulation cell using a lookup table derived from the Integrated Biosphere Simulator [IBIS; @foley_integrated_1996; @liu_critical_2020], a process-based dynamic global vegetation model. We initiated IBIS with minimal vegetation and simulated forward for 110 years using 30-year climate normals for the Hawaiian Islands [@giambelluca_online_2013; @giambelluca_evapotranspiration_2014]. We calibrated IBIS carbon stocks with statewide gridded datasets of soil organic carbon [@ssurgo_2016] and forest aboveground live biomass [@asner_rapid_2016]. We also calibrated gross photosynthesis in IBIS using a Hawaiʻi-specific gridded dataset derived from MODIS satellite imagery [@kimball_evaluating_2017].

@@

```{r fig.align = 'center', out.width = "99%", fig.cap = "Sensitivity of projected changes in total ecosystem carbon storage (a) and mean annual net biome productivity (b) to different rates of carbon dioxide fertilization in the seven main Hawaiian Islands under the RCP 8.5 radiative forcing and high land use scenario. The carbon dioxide fertilization effect (CFE) is the percent change in net primary productivity (NPP) for every 100 ppm increase in atmospheric carbon dioxide. The CFE for all rates is capped at 600 ppm, which is achieved around the year 2060. Solid lines in (a) indicate the mean total ecosystem carbon storage across 30 Monte Carlo realizations for each CFE rate, with shaded areas indicating the minimum and maximum range of Monte Carlo realizations. Solid circles in (b) represent mean annual net biome productivity averaged across all years and Monte Carlo realizations for each CFE rate, with vertical lines indicating the standard deviation of the mean. The dashed horizontal line in (b) represents the boundary between ecosystems acting as a net carbon sink (positive NBP values) and a net carbon source (negative NBP values)."}
knitr::include_graphics("./fig_images/fig6_cfe.png") 
```

@@

This study was funded by the U.S. Geological Survey Biological Carbon Sequestration Program. Thanks to Leonardo Frid and Colin Daniel of ApexRMS for assistance with SyncroSim software and to Nicholas Koch of Forest Solutions, Inc. for information on eucalyptus harvesting in Hawaiʻi. Thanks also to Christian Giardina and Zhiliang Zhu for making this project possible. Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.
